Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-21 cordillera-cycle.tex
Add last corrections from Martin.

@@

In general, the topographic complexity of the North American Cordillera and its
effect on glacial history have inhibited the construction of ice sheet-wide
glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle, and
analyse some of the long-standing questions related to its evolution:

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and MacKenzie Mountains (SMKM). Although the Coast,
Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
MacKenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
\citealp[Fig.~2]{Kleman.etal.2010}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range
\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. A well-developed network of glacial valleys west of
the MacKenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}). However, because it has only been glaciated for a
short fraction
of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
results, this perhaps appears indicate that
the observed landscape pattern originates from multiple glacial cycles and
witnesses an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
west of the MacKenzie Mountains are areas of higher-than-average geothermal
heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
durations of warm-based ice cover for these areas if we were to include
spatially variable geothermal forcing in our Cordilleran ice sheet simulations.

@@

In the North American Cordillera, the presence of lateral melt-water channels
at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
Margold.etal.2014}, and abundant esker systems at low elevation
\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
Margold.etal.2013}
indicate that abundant meltwater was produced over the entire ice sheet surface
during deglaciation. The southern and northern margins of the
Cordilleran ice sheet reached their last glacial maximum extent around
17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
Stroeven.etal.2010, Stroeven.etal.2014}, which we take as a limiting age for
the onset of ice retreat. The timing of final deglaciation is less well
constrained, but recent cosmogenic dates from north-central British Columbia
indicate that a seizable ice cap emanating from the central Coast Mountains or
the Skeena Mountains persisted into the Younger Dryas chronozone, at least
until 12.4\,ka \citep{Margold.etal.2014}.

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Fraser Plateau, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

@@

Although further work is needed to constrain the timing of the late-glacial
readvance, to assess its extents and geographical distribution, and to identify
the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
to note that the simulation driven by the GRIP record produces a late-glacial
readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
corresponding to where it has been identified in the geological record
(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
local readvances, the GRIP-driven simulation shows that larger ice bodies
emanating from the Skeena and MacKenzie mountains may still have existed at the
time of this late-glacial readvance.
In contrast, the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a tightly restricted glacial readvance
(Fig.~\ref{fig:deglac}, right panel).

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelope
\citep[terminology from][]{Kleman.etal.2006}, having been formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Fraser Plateau, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the MacKenzie Mountains.

@@

However, on the Fraser Plateau, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
results could be that the Fraser Plateau lineation system predates
deglaciation, and that deglacial landforms are largely absent. Alternatively,
local geothermal heat associated with volcanic activity on the Fraser Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

@@

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and MacKenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by moisture intruding inland from the
west through a topographic low in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre at the onset of the LGM readvance. As indicated
by persistent, warm-based ice in the model, this ice cap is perhaps associated
with the distinct glacial erosional imprint on the landscape of the Skeena
Mountains.

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and MacKenzie mountains (SMKM),
           the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Fraser Plateau (FP) and the
           Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:locmap}
\end{figure}

>>>>

% Todo
% change Fraser Plateau to Interior Plateau

@@

In general, the topographic complexity of the North American Cordillera and its
effect on glacial history have inhibited the reconstruction of ice sheet-wide
glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle, and
analyse some of the long-standing questions related to its evolution:

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie Mountains (SMKM). Although the Coast,
Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
\citealp[Fig.~2]{Kleman.etal.2010}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range
\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. A well-developed network of glacial valleys west of
the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}). However, because it has only been glaciated for a
short fraction
of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
results, this perhaps appears indicate that
the observed landscape pattern originates from multiple glacial cycles and
witnesses an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
west of the Mackenzie Mountains are areas of higher-than-average geothermal
heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
durations of warm-based ice cover for these areas if we were to include
spatially variable geothermal forcing in our Cordilleran ice sheet simulations.

@@

In the North American Cordillera, the presence of lateral meltwater channels
at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
Margold.etal.2014}, and abundant esker systems at low elevation
\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
Margold.etal.2013}
indicate that abundant meltwater was produced over the entire ice sheet surface
during deglaciation. The southern and northern margins of the
Cordilleran ice sheet reached their last glacial maximum extent around
17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
Stroeven.etal.2010, Stroeven.etal.2014}, which we take as a limiting age for
the onset of ice retreat. The timing of final deglaciation is less well
constrained, but recent cosmogenic dates from north-central British Columbia
indicate that a seizable ice cap emanating from the central Coast Mountains or
the Skeena Mountains persisted into the Younger Dryas chronozone, at least
until 12.4\,ka \citep{Margold.etal.2014}.

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Fraser Plateau, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

@@

Although further work is needed to constrain the timing of the late-glacial
readvance, to assess its extents and geographical distribution, and to identify
the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
to note that the simulation driven by the GRIP record produces a late-glacial
readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
corresponding to where it has been identified in the geological record
(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
local readvances, the GRIP-driven simulation shows that larger ice bodies
emanating from the Skeena and Mackenzie mountains may still have existed at the
time of this late-glacial readvance.
In contrast, the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a tightly restricted glacial readvance
(Fig.~\ref{fig:deglac}, right panel).

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelope
\citep[terminology from][]{Kleman.etal.2006}, having been formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Fraser Plateau, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the Mackenzie Mountains.

@@

However, on the Fraser Plateau, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
results could be that the Fraser Plateau lineation system predates
deglaciation, as perhaps indicated by some eskers that appear incompatible with
these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
local geothermal heat associated with volcanic activity on the Fraser Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

@@

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by moisture intruding inland from the
west through a topographic low in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre at the onset of the LGM readvance. As indicated
by persistent, warm-based ice in the model, this ice cap is perhaps associated
with the distinct glacial erosional imprint on the landscape of the Skeena
Mountains.

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains (SMKM),
           the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Fraser Plateau (FP) and the
           Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:locmap}
\end{figure}
